---
title: "CodeDepends: Static analysis and dependency detection for R code"
author: "Gabriel Becker"
output:
  rmarkdown::html_vignette:
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{CodeDependsIntro}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# Introduction

The CodeDepends package provides a flexible framework for statically
analyzing R code (i.e., without evaluating it). It also contains
higher-level functionality for: detecting dependencies between R code
blocks or expressions, "tree-shaking" (pruning a script down to only
the expressions necessary to evaluate a given expression), plotting
variable usage timelines, and more.

# The workhorses: readScript and getInputs

The primary functions to perform basic code analysis are `readScript`
which reads in R scripts of various forms (including .R and .Rmd
files), and `getInputs` which performs the low-level code-analysis.

The `readScript` function returns a `Script` object (essentially a
list of `ScriptNodes` representing the top-level expressions in the
script). This can then be passed to the `getInputs` which, in that
case, returns a `ScriptInfo` object, which can be thought of as a list
of `ScriptNodeInfo` objects representing information about those
top-level expressions.

R expressions can also be passed directly to `getInputs`, which
returns a single `ScriptNodeInfo` object in that case. While in
practice users will generally call `getInputs` on entire scripts,
passing expressions directly is useful for testing and illustration.

As stated above, `ScriptNodeInfo` objects are the units of information
about single expressions being analyzed, and collect various
information extracted from examining the expression itself:

```{r scriptnodeinfo}
library(CodeDepends)
getInputs(quote(x <- y + rnorm(10, sd = z)))

```

As we can see, the information includes the any string literals used
in the expression, split into file and non-file strings based on
whether the string appears to point to an existing path at analysis
time with respect to the `basedir` argument (which defaults to the
current directory). It also contains any libraries loaded by the code
(via `library`, `require`, or `requireNamespace` calls).

Next is are the inputs and outputs of the expression, which are the
variables used by the expression and created by the expression (via
assignment), respectively. By default, these lists will not include
symbols used in ways that mean they are non-standardly evaluated
(e.g., within the construction of a `ggplot2` plot object). These
non-standard evaluation variables are collected separately (as
nsevalVars).

Variables whose values are updated (ie ones who are assigned new values which depend on their existing value) are collected separately. These updates can take a large number of forms, including:

```{r updateexprs, eval=FALSE}
x = x + 5
rownames(x) = 5
x[1:3] = 5
x  = lapply(1:5, function(i) x[i]^2)
x$y = 5
```

In all of the above cases, the variable `x` will be listed in both the
`updated` and `inputs` categories, but *NOT* in the `outputs`
category.

Next are the functions which were called by the expression. These
include those invoked as funtionals, e.g. via the `apply` family or
`mutate_*` and `summarize_*` families. We note here that the functions
list is actually a `logical` vector, indicating whether the function
was locally defined within the script (`TRUE`), defined within a
package (`FALSE`), or unkown (`NA`). The names of the vector indicate
the names of the functions. Currently, functions will always be
unknown if a single expression is analyzed directly. Function
provenance detection is only applied to full scripts.

Finally, the list of removed variables, side-effects `CodeDepends` is
able to detect, and a copy of the code complete the list of
information extracted.

## Symbols within formulas

Symbols within formulas are treated specially when analyzing code, based on the `formulaInputs` argument to `getInputs`. If `FALSE` (the default), they are assumed to evaluated nonstandardly (e.g., in the context of a `data.frame`), if `TRUE`, they are counted as standard inputs. Currently there is no capacity for mixing these behaviors within a single call to `getInputs`. 

# Input collectors, function handlers, and customization

The `getInputs` function accepts a `collector` argument, which
essentially specifies a state tracker to be used when walking the code
to collect inputs, functions called, etc.

For largely historical reasons, input collectors are roughly defined as the output from the `inputCollector` constructor, rather than as a more formal class.

When creating an input collector, various behavior can be customized,
primarily in the form of \function handlers\ which specify behavior
when analyzing calls to specific functions. This is, for example, how
`CodeDepends` knows that some arguments within certain functions are
non-standardly evaluated. CodeDepends ships with a robust set of
default handlers, but these can be overridden or supplemented with
custom handlers by specifying them when constructing the collector,
either via the `...` arguments or as list. In both cases, the names
are the names of the function the handler should be used on.

```{r custhandler}

col = inputCollector(library = function(e, collector, ...) {
    print(paste("Hello", asVarName(e)))
    defaultFuncHandlers$library(e, collector, ...)
})
getInputs(quote(library(CodeDepends)), collector = col)
```

`inputCollector` also accepts arguments which control what is counted
as an input when processing expressions. The `inclPrevOutput` argument
specifies whether output variables should be included as inputs to
subsequent expressions when processing multiple expressions as an
single block (e.g., when they are wrapped in `{}`). The
`checkLibrarySymbols` and `funcsAsInputs` arguments control how
symbols which appear to be resolved within libraries, and functions
which are called are handled, respectively. The default behavior is
for all of these to be `FALSE`.

# Dependency detection and script visualization

`CodeDepends` can visualize code in various ways.

## Variable dependency graphs
We can create the variable graph of dependnecies between variables,
via the `makeVariableGraph` function:

```{r variablegraph}

 f = system.file("samples", "results-multi.R", package = "CodeDepends")
 sc = readScript(f)
 g = makeVariableGraph( info = getInputs(sc))
 if(require(Rgraphviz))
   plot(g)
```

## call graphs

We can also create call graphs for functions or entire packages:

```{r callgraphs}
  gg = makeCallGraph("package:CodeDepends")
  if(require(Rgraphviz)) {
     gg = layoutGraph(gg, layoutType = "circo")
     graph.par(list(nodes = list(fontsize=55)))
     renderGraph(gg) ## could also call plot directly
  } 
```

## Variable definitions timelines

Finally we can display timelines for when variables are defined,
redefined, and used:

```{r timelines}
f = system.file("samples", "results-multi.R", package = "CodeDepends")
sc = readScript(f)
dtm = getDetailedTimelines(sc, getInputs(sc))
plot(dtm)

 # A big/long function
info = getInputs(arima0)
dtm = getDetailedTimelines(info = info)
plot(dtm, var.cex = .7, mar = 4, srt = 30)
```
